---
title: "The \tt{mpcmp} Package for Mean-Parametrizied Conway-Maxwell-Poisson Regression"
author: "Thomas Fung, Aya Alwan, Justin Wishart and Alan Huang"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The \tt{mpcmp} Package for Mean-Parametrizied Conway-Maxwell-Poisson Regression}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

We review the theory and application of a mean-parametrized Conway-Maxwell-Poisson generalzied linear models for dispersed count data and describe the estimation of these models using the R package mpcmp. Prediction, inference, diagnostic and graphical methods are also illustrated by several examples. 

## Introduction 

Conway–Maxwell–Poisson (COM-Poisson) distributions have seen a recent resurgence in popularity for the analysis of dispersed counts (e.g., Shmueli et al., 2005; Lord et al., 2008, 2010; Sellers and Shmueli, 2010). Key features of COM-Poisson distributions include the ability to handle both overdispersion and underdispersion, containing the classical Poisson distribution as a special case, and being a continuous bridge between other classical distributions such as the geometric and Bernoulli distributions. COM-Poisson distributions are also full probability models, making them particularly useful for predictions and estimation of event probabilities. 


Vignettes are long form documentation commonly included in packages. Because they are part of the distribution of the package, they need to be as compact as possible. The `html_vignette` output type provides a custom style sheet (and tweaks some options) to ensure that the resulting html is as small as possible. The `html_vignette` format:

- Never uses retina figures
- Has a smaller default figure size
- Uses a custom CSS stylesheet instead of the default Twitter Bootstrap style

## Vignette Info

Note the various macros within the `vignette` section of the metadata block above. These are required in order to instruct R how to build the vignette. Note that you should change the `title` field and the `\VignetteIndexEntry` to match the title of your vignette.

## Styles

The `html_vignette` template includes a basic CSS theme. To override this theme you can specify your own CSS in the document metadata as follows:

    output: 
      rmarkdown::html_vignette:
        css: mystyles.css

## Figures

The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, fig.show='hold'}
plot(1:10)
plot(10:1)
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.

## More Examples

You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(mtcars, 10))
```

Also a quote using `>`:

> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))
